DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-19 manuscript/manuscript-reservoir.tex
subsection 3.2

is the correction of the displacements considering a semi-space (image nucleus solution).

@@

For convenience, the $x-$, $y-$ and $z-$components of the vectors ${\bf{u_{1}}}_{i}$ (equation \ref{eq:nucleus-u1}) and ${\bf{u_{2}}}_{i}$ (equation \ref{eq:nucleus-u2}) 

@@

We parameterize the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has a constant pressure contrast $\Delta p_{j}$; however, pressure variations among different prisms are allowed. 
To calculate the displacement field produced by the $j$th prism
at the $i$th coordinates  $(x_i$, $y_i, z_i)$, we integrate the solution deduced for a single nucleus of strain (equation \ref{eq:displacement_ui})
over its volume and obtain
{\bf{\tilde{u}}}_{ij} = 
\iiint\limits_{{v_{1}}_{j}}
{\bf{u_{1}}}_{ij} \: \:  dv_{j}^{\prime}
\: + \:
\iiint\limits_{{v_{2}}_{j}}
{\bf{u_{2}}}_{ij} \:\:  dv_{j}^{\prime} \: ,
\label{eq:nucleus-ui_jth_prism}
where both integrals are conducted with respect to the variables $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$, but the first is conducted over the volume ${v_{1}}_{j}$ of the $j$th prism and the second is conducted over the volume 
${v_{2}}_{j}$ of a different prism symmetrically positioned above the free surface 
and conveniently called $j$th "image prism".
Volume ${v_{1}}_{j}$ is defined by ${x_1}_{j}$, ${x_2}_{j}$, ${y_1}_{j}$, ${y_2}_{j}$,
${z_1}_{j}$, and ${z_2}_{j}$, which represent, respectively, the south, north, west, 
east, top, and bottom borders of the $j$th prism.
Volume ${v_{2}}_{j}$ of the $j$th image prism is defined in a similar way, but with 
top and bottom borders given by $z_1 - 2 \, z_c$ and $z_2 - 2 \, z_c$, where 
$z_c = \frac{1}{2}(z_1 + z_2)$ is the center depth of the $j$th prism.
By substituting equations \ref{eq:nucleus-u1-vector} and \ref{eq:nucleus-u2-vector} into equation \ref{eq:nucleus-ui_jth_prism}, we obtain
{\tilde{u}}_{{ij}_\alpha} = \frac{A  \: (1 + \nu)}{E} 
\: \Delta p_{j}  \: \left[ \iiint\limits_{{v_{1}}_{j}} 
\frac{\partial }{\partial \alpha} {\frac{1}{{R_1}_{ij}}} \:  dv_{j}^{\prime} 
\: + \: 
(3  - 4 \nu)  \iiint\limits_{{v_{2}}_{j}}
\: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \iiint\limits_{{v_{2}}_{j}}
\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_{j}^{\prime} 
\right] \: , 
\label{eq:u_til_alpha}
where $\alpha = x, y$, and
{\tilde{u}}_{{ij}_z} = \frac{A  \: (1 + \nu)}{E} 
\: \Delta p_{j}  \: \left[ \iiint\limits_{{v_{1}}_{j}}
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} \:  dv_j 
\: - \: (3  - 4 \nu)  \iiint\limits_{{v_{2}}_{j}}
\: \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \iiint\limits_{{v_{2}}_{j}}
\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_j 
\right] \: .
\label{eq:u_til_z}
In the right-hand side of equations \ref{eq:u_til_alpha} and \ref{eq:u_til_z}, the three integrals have the same form of derivatives of the gravitational potential produced by the 
$j$th prism. 
The first integral corresponds to the $\alpha-$component of the gravitational attraction 
produced by the $j$th prism.
The second and third integrals correspond, respectively, to the $\alpha-$component of the gravitational attraction and to the $\alpha \:z-$component of the gravitational 
gradient tensor produced by the $j$th image prism. 
The similarity between the displacement fields due to a volume source in a half-space and the gravitational field allows the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism.  

@@

y \ln(z + {R_1}) + z \ln(y + {R_1}) -  x  \tan^{-1} \Bigg( \frac{yz}{x \:{R_1}} \Bigg) 

@@

x \ln(z + {R_1}) + z \ln(x + {R_1}) -  y  \tan^{-1} \Bigg( \frac{xz}{y \:{R_1}} 

@@

x \ln(y + {R_1}) + y \ln(x + {R_1}) -  z  \tan^{-1} \Bigg( \frac{xy}{z \:{R_1}} \Bigg) 

@@

$(x_i, y_i, z_i)$, ${R_1} = \sqrt{x^{2} + y^{2} + z^{2}}$, and

@@

\label{eq:Nagy_solutions_1}

@@

PAREI AQUI

For simplicity,  we omit the subscripts $i$ and $j$ in equations \ref{dx1} $-$ \ref{dz1}; hence, the variables $x$, $y$ and $z$  are relative coordinates of the $i$th point of the displacement referred to the coordinates to the corner of the $j$th prism modeling the reservoir, i.e.,  $x = x_j - x_i $, $\: y = y_j - y_i $, $\: z = z_j - z_i $ and 
$\: {R_1} = \sqrt{x^{2} + y^{2} + z^{2}}$.
In equations \ref{dx1} $-$ \ref{dz1}, the limits of the integrals represent the borders of the $j$th prism modeling the reservoir in the following way: $x_1$ and $x_2$ are their south and north borders; $y_1$ and  $y_2$ are their  west and east borders;  and 
$z_1$ and $z_2$ are their depths to the top and bottom.
\cite{Nagyetal2000} and (\citeyear{Nagyetal2002}) provided limit values of integrals shown in equations \ref{dx1} $-$ \ref{dz1} when the computation point coincides with the corner of the prism.

The second and third integrals in the right-hand side of equations  \ref{eq:u_til_alpha}  and \ref{eq:u_til_z} are related with the correction of the displacements considering  a semi-space (image nucleus solution).
These integrals depend on the distance ${R_2}_{ij}$ from the $i$th coordinate point of the displacement to the $j$th image nucleus.
Likewise, $\: {R_2} = \sqrt{x^{2} + y^{2} + z^{2}}$; however, the variable $z$  that represents the relative coordinate of the $i$th point of the displacement referred to the coordinates to the corner of the $j$th image nucleus is given by
\begin{equation}
z = z_j - z_i - 2 z_c
\label{z_image}
\end{equation}
where  $z_c$ is the $z-$coordinate of the $j$th image nucleus
\begin{equation}
z_c = 0.5 (z_1 + z_2)
\label{z_c}
\end{equation}

The second integral in the right-hand side of  equations  \ref{eq:u_til_alpha} and  \ref{eq:u_til_z} is the first derivatives of ${\frac{1}{{R_2}_{ij}}}$ with respect to $x$, $y$ and $z$.
These derivatives are equal to equations \ref{dx1} $-$ \ref{dz1}, the only difference is the variable $z$  that is given by equations \ref{z_image}  and \ref{z_c}.

The third integral in the right-hand side of  equations  \ref{eq:u_til_alpha}  and  \ref{eq:u_til_z} is the second derivatives of ${\frac{1}{{R_2}_{ij}}}$ with respect to $xz$, $yz$ and $zz$ which are, respectively, given by the following closed expressions: 
\begin{equation}
\int\int\limits_{v_j}\int 
\frac{\partial^{2}  }{\partial x \partial z} {\frac{1}{{R_2}_{ij}}}  dv_j = \Bigg|\Bigg|\Bigg| 
\ln(y + {R_2})
\Bigg|_{x_1}^{x_2} \Bigg|_{y_1}^{y_2} \Bigg|_{z_1}^{z_2}
\label{dx2}
\end{equation}
\begin{equation}
\int\int\limits_{v_j}\int
\frac{\partial^{2}  }{\partial y \partial z} {\frac{1}{{R_2}_{ij}}}  dv_j =
\Bigg|\Bigg|\Bigg|
\ln(x + {R_2})
\Bigg|_{x_1}^{x_2} \Bigg|_{y_1}^{y_2} \Bigg|_{z_1}^{z_2}
\label{dy2}
\end{equation}
\begin{equation}
\int\int\limits_{v_j}\int
\frac{\partial^{2}  }{\partial z \partial z} {\frac{1}{{R_2}_{ij}}}  dv_j =
\Bigg|\Bigg|\Bigg|
 -   \tan^{-1} \Bigg( \frac{xy}{z \:{R_2}} \Bigg)
\Bigg|_{x_1}^{x_2} \Bigg|_{y_1}^{y_2} \Bigg|_{z_1}^{z_2}, 
\label{dz2}
\end{equation}
where the variable $z$ is given by equations \ref{z_image}  and \ref{z_c}.

% Displacement produced by the whole reservoir 
Finally, the displacement field at the point $(x_i, y_i, z_i)$ due to the pressure change in the whole reservoir is defined as the sum of the displacements 
${\bf{u}}_{ij}$ (equation \ref{eq:nucleus-ui_jth_prism}) yielded by each prism 
with constant pressure $\Delta p_{j}$:
\begin{equation}
{\bf{\tilde{u}}}_{ij} = \sum_{j=1}^{M} {\bf{u}}_{ij} \: ,
\label{eq:nucleus-ui_M_prisms}
\end{equation}
where $M$ is the number of prisms setting up the reservoir model.
The horizontal displacement field at the $i$th coordinates ($x_i$, $y_i$  and $z_i$) due to the pressure change in the whole reservoir is calculated by 
\begin{equation}
{\tilde{u}}_{{i}_h} = \sqrt{ {\tilde{u}}_{{i}_x}^{2}  +  {\tilde{u}}_{{i}_y}^{2} }  
\label{eq:horizontal_displacement}
\end{equation}
where ${\tilde{u}}_{{i}_x}$ and ${\tilde{u}}_{{i}_y}$ are the $x-$ and $y-$ components of the displacement field at the $i$th coordinates given by equation \ref{eq:u_til_alpha}, with $\alpha = x $ and $y$.

>>>>

is the correction of the displacement considering a semi-space (image nucleus solution).

@@

The $x-$, $y-$ and $z-$components of the vectors ${\bf{u_{1}}}_{ij}$ (equation \ref{eq:nucleus-u1}) and ${\bf{u_{2}}}_{ij}$ (equation \ref{eq:nucleus-u2}) 

@@

We parameterize the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has a constant pressure contrast $\Delta p_{j}$; however, pressure variations among different prisms are allowed. 
To calculate the displacement field produced by the $j$th prism
at the $i$th coordinates  $(x_i$, $y_i, z_i)$, we integrate the solution deduced for a single nucleus of strain (equation \ref{eq:displacement_ui})
over its volume and obtain
\begin{equation}
{\bf{\tilde{u}}}_{ij} = 
\iiint\limits_{{v_{1}}_{j}}
{\bf{u_{1}}}_{ij} \: \:  dv_{j}^{\prime}
\: + \:
\iiint\limits_{{v_{2}}_{j}}
{\bf{u_{2}}}_{ij} \:\:  dv_{j}^{\prime} \: ,
\label{eq:nucleus-ui_jth_prism}
\end{equation}
where both integrals are conducted with respect to the variables 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. The first integral in equation 
\ref{eq:nucleus-ui_jth_prism} is conducted over the volume ${v_{1}}_{j}$ of the $j$th
prism and the second is conducted over the volume 
${v_{2}}_{j}$ of a different prism symmetrically positioned above the free surface 
and conveniently called $j$th "image prism".
Volume ${v_{1}}_{j}$ is defined by ${x_1}_{j}$, ${x_2}_{j}$, ${y_1}_{j}$, ${y_2}_{j}$,
${z_1}_{j}$, and ${z_2}_{j}$, which represent, respectively, the south, north, west, 
east, top, and bottom borders of the $j$th prism.
Volume ${v_{2}}_{j}$ of the $j$th image prism is defined in a similar way, but with 
top and bottom given by $z_1 - 2 \, z_c$ and $z_2 - 2 \, z_c$, where 
$z_c = \frac{1}{2}(z_1 + z_2)$ is the center depth of the $j$th prism.
By substituting equations \ref{eq:nucleus-u1-vector} and \ref{eq:nucleus-u2-vector} 
into equation \ref{eq:nucleus-ui_jth_prism}, we obtain
\begin{equation}
{\tilde{u}}_{{ij}_\alpha} = \frac{A  \: (1 + \nu)}{E} 
\: \Delta p_{j}  \: \left[ \iiint\limits_{{v_{1}}_{j}} 
\frac{\partial }{\partial \alpha} {\frac{1}{{R_1}_{ij}}} \:  dv_{j}^{\prime} 
\: + \: 
(3  - 4 \nu)  \iiint\limits_{{v_{2}}_{j}}
\: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \iiint\limits_{{v_{2}}_{j}}
\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_{j}^{\prime} 
\right] \: , 
\label{eq:u_til_alpha}
\end{equation}
where $\alpha = x, y$, and
\begin{equation}
{\tilde{u}}_{{ij}_z} = \frac{A  \: (1 + \nu)}{E} 
\: \Delta p_{j}  \: \left[ \iiint\limits_{{v_{1}}_{j}}
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} \:  dv_j 
\: - \: (3  - 4 \nu)  \iiint\limits_{{v_{2}}_{j}}
\: \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \iiint\limits_{{v_{2}}_{j}}
\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_j 
\right] \: .
\label{eq:u_til_z}
\end{equation}
In the right-hand side of equations \ref{eq:u_til_alpha} and \ref{eq:u_til_z}, the three integrals have the same form of derivatives of the gravitational potential produced by the 
$j$th prism. 
The first integral corresponds to the $\alpha-$component of the gravitational attraction 
produced by the $j$th prism.
The second and third integrals correspond, respectively, to the $\alpha-$component of the gravitational attraction and to the $\alpha \:z-$component of the gravitational 
gradient tensor produced by the $j$th image prism. 
The similarity between the displacement fields due to a volume source in a half-space and the gravitational field allows the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism.

@@

y \ln(z + R) + z \ln(y + R) -  x  \tan^{-1} \Bigg( \frac{yz}{x \: R} \Bigg) 

@@

x \ln(z + R) + z \ln(x + R) -  y  \tan^{-1} \Bigg( \frac{xz}{y \: R} 

@@

x \ln(y + R) + y \ln(x + R) -  z  \tan^{-1} \Bigg( \frac{xy}{z \: R} \Bigg) 

@@

$(x_i, y_i, z_i)$, $R = \sqrt{x^{2} + y^{2} + z^{2}}$, and

@@

\label{eq:Nagy_limits_1}

@@

The remaining integrals depending on first and second derivatives 
of ${\frac{1}{{R_2}_{ij}}}$ in the right-hand side of equations 
\ref{eq:u_til_alpha} and \ref{eq:u_til_z} are conducted over the volume 
${v_{2}}_{j}$ of the $j$th image prism and have the following closed solutions 
\int\limits_{{x_{1}}_{j}}^{{x_{2}}_{j}} \int\limits_{{y_{1}}_{j}}^{{y_{2}}_{j}} \int\limits_{{z_{1}}_{j} - 2 \, z_c}^{{z_{2}}_{j} - 2 \, z_c}
\frac{\partial }{\partial x} {\frac{1}{{R_2}_{ij}}} dv_{j}^{\prime} =
y \ln(z + R) + z \ln(y + R) -  x  \tan^{-1} \Bigg( \frac{yz}{x \: R} \Bigg) 
\label{dx2}
\int\limits_{{x_{1}}_{j}}^{{x_{2}}_{j}} \int\limits_{{y_{1}}_{j}}^{{y_{2}}_{j}} \int\limits_{{z_{1}}_{j} - 2 \, z_c}^{{z_{2}}_{j} - 2 \, z_c}
\frac{\partial }{\partial y} {\frac{1}{{R_2}_{ij}}} dv_{j}^{\prime} =
x \ln(z + R) + z \ln(x + R) -  y  \tan^{-1} \Bigg( \frac{xz}{y \: R} 
\label{dy2}
\int\limits_{{x_{1}}_{j}}^{{x_{2}}_{j}} \int\limits_{{y_{1}}_{j}}^{{y_{2}}_{j}} \int\limits_{{z_{1}}_{j} - 2 \, z_c}^{{z_{2}}_{j} - 2 \, z_c}
\frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}} dv_{j}^{\prime} =
x \ln(y + R) + y \ln(x + R) -  z  \tan^{-1} \Bigg( \frac{xy}{z \: R} \Bigg) 
\label{dz2}
\int\limits_{{x_{1}}_{j}}^{{x_{2}}_{j}} \int\limits_{{y_{1}}_{j}}^{{y_{2}}_{j}} \int\limits_{{z_{1}}_{j} - 2 \, z_c}^{{z_{2}}_{j} - 2 \, z_c}
\frac{\partial^{2}  }{\partial x \partial z} {\frac{1}{{R_2}_{ij}}}  dv_j = \Bigg|\Bigg|\Bigg| 
\ln(y + R)
\label{dxz2}
\int\limits_{{x_{1}}_{j}}^{{x_{2}}_{j}} \int\limits_{{y_{1}}_{j}}^{{y_{2}}_{j}} \int\limits_{{z_{1}}_{j} - 2 \, z_c}^{{z_{2}}_{j} - 2 \, z_c}
\frac{\partial^{2}  }{\partial y \partial z} {\frac{1}{{R_2}_{ij}}}  dv_j =
\ln(x + {R_2})
\label{dyz2}
\int\limits_{{x_{1}}_{j}}^{{x_{2}}_{j}} \int\limits_{{y_{1}}_{j}}^{{y_{2}}_{j}} \int\limits_{{z_{1}}_{j} - 2 \, z_c}^{{z_{2}}_{j} - 2 \, z_c}
\frac{\partial^{2}  }{\partial z \partial z} {\frac{1}{{R_2}_{ij}}}  dv_j =
-   \tan^{-1} \Bigg( \frac{xy}{z \: R} \Bigg)
\Bigg|_{{X_1}_{j}}^{{X_2}_{j}} \Bigg|_{{Y_1}_{j}}^{{Y_2}_{j}} \Bigg|_{{Z_1}_{j}}^{{Z_2}_{j}} \quad .
\label{dzz2}
In these integrals related to the $j$th image prism 
(equations \ref{dx2}--\ref{dzz2}), the integration limits along the $z$ direction
are given by
{Z_1}_{j} &= z_i - {z_1}_{j} + 2 \, z_c \\
{Z_2}_{j} &= z_i - {z_2}_{j} + 2 \, z_c
\end{array} \quad ,
\label{eq:Nagy_limits_2}
where $z_c = \frac{1}{2}(z_1 + z_2)$ is the center depth of the $j$th prism. 
The remaining limits along $x$ and $y$ directions are the same defined by equation 
\ref{eq:Nagy_limits_1}.

% Displacement produced by the whole reservoir 
Finally, the total displacement vector at the point $(x_i, y_i, z_i)$ due to the
pressure change in the whole reservoir is defined as the sum of the displacements 
${\bf{u}}_{ij}$ (equation \ref{eq:nucleus-ui_jth_prism}) yielded by each prism 
with constant pressure $\Delta p_{j}$:
\begin{equation}
{\bf{\tilde{u}}}_{i} = \sum_{j=1}^{M} {\bf{\tilde{u}}}_{ij} \: ,
\label{eq:nucleus-ui_M_prisms}
\end{equation}
where $M$ is the number of prisms setting up the reservoir model.
The horizontal component of the total displacement vector ${\bf{\tilde{u}}}_{i}$ 
(equation \ref{eq:nucleus-ui_M_prisms}) is calculated by 
\begin{equation}
{\tilde{u}}_{{i}_h} = \sqrt{ {\tilde{u}}_{{i}_x}^{2}  +  {\tilde{u}}_{{i}_y}^{2} } 
\: ,
\label{eq:horizontal_displacement}
\end{equation}
where ${\tilde{u}}_{{i}_x}$ and ${\tilde{u}}_{{i}_y}$ are the $x-$ and $y-$ components.
